
---
title: Statistical Analysis of Gene Expression Data
author: David Winter
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    pdf_document:
        latex_engine: xelatex
monofont : FreeMono
mainfont: "Linux Libertine O"
---

##Introduction

This document describes and provides the statistical analyses used to examine
gene expression data from _Epichloë festucae_ strain FL1. The analyses are
based on a data set that contains data derrrived from a number of analyses

```{r}
expr_data <- read.csv("expression_summary.csv", stringsAsFactors=FALSE)
knitr::kable(expr_data[1:5, c("gene","log2FoldChange", "Fl1_specific", "dAT", "tad")])
```

## Do TADs predict gene expression changes?

We first fit a linear model of change in gene expression is predicted by
membership of a TAD.  To do this tidyly we produce a function that parsed model
ouputs.

```{r}

make_mod_tab <- function(mod, fdr=0.1){
    mod_table <- as.data.frame(coef(summary(mod))[-1,])
    mod_table$chrom <- sapply(strsplit(rownames(mod_table), "_"), "[[", 2)
    mod_table$bin <- sapply(strsplit(rownames(mod_table), "_"), "[[", 3)
    mod_table <- mod_table[order(mod_table$chrom, as.numeric(mod_table$bin)),]
    names(mod_table) <- c("est.", "std.error", "t-val", "p-val", "chrom", "tad")
    mod_table$padj <-  p.adjust(mod_table$`p-val`, method="fdr")
    mod_table$sig <- mod_table$padj <= fdr
    mod_table
}
```

Then fit both the full model (with TAD predicting change in gene expressoin) and
a null model. These models are compared by AIC.

```{r}
tad_data <- expr_data[!is.na(expr_data$tad),]
mod <- lm(log2FoldChange ~ tad, data=tad_data)
null <- lm(log2FoldChange ~ 1, data=tad_data)
AIC(mod, null)

```

As the model with TADs fits subtantially better than the mdoel without them, we
will plot the estimated effect-size for each TAD (Figure 10a in the manuscript).

```{r}
library(ggplot2)
library(ggrepel)

res <- make_mod_tab(mod)
res$order <- 1:nrow(res)
p <- ggplot(res, aes(x=order, y=est., ymax=est. + 2 * std.error, ymin=est. - 2 * std.error, colour=chrom))
p + geom_hline(yintercept=0) + geom_errorbar(width=0) + 
    geom_point(size=2) + 
    scale_colour_brewer(palette="Dark2") +
    scale_y_continuous("log2 Fold change", limits=c(-6,6)) +
    geom_label_repel(data=res[ res$sig,], aes(x=order, y=est., label=paste(chrom, tad, sep="_")))

```

## Are secondary metabolite and other important genes close to AT rich regions?

We then consider the possibility that AT-rich regions are associated with
particular classes of genes that may be important in regulating the symbiosis
between _Epichloe_ and its host.

Because we do not expect a linear relatinship between the distance between an
AT-rich region and any of these gene classes we first distcritize the "distance
from AT" data into give non-overlapping bins.

```{r}
at_classes <- c(-Inf, 0, 300, 1000, 5000, Inf)
expr_data$dAT_cut <- cut(expr_data$dAT, at_classes, 
                         ordered_result=TRUE, 
                         labels = c("In", "[0,300bp)", "[300bp,1.kbp)","[1kbp,5kbp)", ">=5kbp")
)

```

We tried two different approaches to testing for association between AT-rich
regions and "ordered probit regression", which would accomidate the orderd
nature of distance-classes and logistic regression in which proximity to an
AT-rich region was treated as binary variable (with values >= 5kb being
considred distant and < 5kb close). In practice, both analyses produced similar
results and the logistic regression is the easiest to interpret biologically, so
we only report this result.


```{r}
d_mod  <- MASS::polr(dAT_cut ~ Fl1_specific + upreg + log(dTelo) + secreted + metab, data=expr_data, Hess=TRUE, method="probit")
p_mod <- glm(dAT < 5000 ~ Fl1_specific + upreg + (dTelo < 5e4) + secreted + metab, data=expr_data, family="binomial")
summary(p_mod)
```

We nevertheless visualise the effects across all distance bins.

```{r}
ggplot(expr_data, aes(dAT_cut, fill=Fl1_specific)) + geom_bar(position="fill", colour="black") + scale_fill_brewer()

predictors <- aggregate(cbind(Fl1_specific, metab, secreted, upreg) ~ dAT_cut, data=expr_data, FUN=sum)
molten <- melt(predictors)
molten$n <- table(expr_data$dAT_cut)
ggplot(molten, aes(dAT_cut, value/n, group=variable, colour=variable, label=variable)) + 
    geom_point(size=2) + geom_line(size=1) + 
    scale_colour_brewer(palette="Dark2")

```
